A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-01-26 analysis/paper/paper.Rmd
Figuring out how to plot Brunlanes dates

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The best point estimate for the distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m. However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to m below the site location. This is constrained further when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. When applying the method of shoreline dating to Stone Age sites in the region, it could thus be recommended, given no other information, to allow for a potential deviation between site and shoreline from `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))` in vertical distance. This can be reduced to `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m for Mesolithic sites. The sites treated here have been given a shoreline date using these values in Figure. However,      
An exponential function has been fit to the distribution. While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship. The accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero.

@@

```{r shoredate_sites, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobases and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline. The shoreline dates in gray are plotted against the oldest radiocarbon-informed dates for each site in black. The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}

@@

>>>>

Accepting that shoreline dating appears to loose utility around the transition to the Late Neolithic, the results for the second row of Figure \@ref(fig:results) is given again in Figure \@ref(fig:results2), excluding all simulation results younger than 2500 BCE. Furthermore, all negative values have here been set to zero, under the assumption that these result from uncertainty or errors in the data, and not actual site locations. The best point estimate for the distance between site and shoreline for the pre-Late Neolithic is, following from the skewed distribution, given by the median at `r round(median(distances2$vertdist))`m. However, 95% of the distance values fall within the range `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))`m. That is, for 95% of the cases, the shoreline was situated on or down to m below the site location. This is constrained further when only considering the Mesolithic dates, with a median offset of `r round(median(distances3$vertdist))`m and with 95% of distance measures within the range `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m. When applying the method of shoreline dating to Stone Age sites in the region, it could thus be recommended, given no other information, to allow for a potential deviation between site and shoreline from `r round(quantile(distances2$vertdist, 0))`--`r round(quantile(distances2$vertdist, 0.95))` in vertical distance. This can be reduced to `r round(quantile(distances3$vertdist, 0))`--`r round(quantile(distances3$vertdist, 0.95))`m for Mesolithic sites.

An exponential function has been fit to the distribution. While it is clear that this does not perfectly match the data, it would seem reasonable from a theoretical perspective to assume that a process of exponential decay explains this relationship. The accumulation of distances-values on the 0m mark likely follow from forcing negative values to zero, from the resolution of the spatial data, and from defining intersecting sea- and site polygon as having a distance of zero. The sites treated here have been given a shoreline date using these values in Figure Figure \@ref(fig:shoredate).

@@

```{r shoredate, eval = TRUE, echo = FALSE, fig.cap = "The result of shoreline dating the analysed sites by mean elevation of the site polygon, accounting for distance to closest isobase and employing the exponential decay function identified in Figure 7 for the distance between sites and shoreline. The shoreline dates in gray are plotted against the oldest radiocarbon-informed phase for each site in black. The red line indicates the 95% probability range for the shoreline date.", out.width = "525px", fig.align="center"}

@@

# Shoreline dating the pioneer sites from Brunlanes

The pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway. They have a distinct Early Mesolithic artefact inventory and are situated in an extremely steep area of the landscape where it would be difficult to envision any use of the sites after the sea retreated any significant distance from their location. Not least following from the fact that no pre-Boreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of these earliest sites are of fundamental importance for understanding the pioneer settlement and intial colonisation of the Scandinavian peninsula. In the original publication of the sites, Jaksland provides a thorough discussion of shoreline dating in general and as used for the dating of the Brunlanes sites specifically. The result of comparing his results against the ones achieved using the above-outlined approach are given in Figure. These are largely in agreement. The sites are located more or less on the isobase of the displacement curve for Larvik and so neglible adjustment was needed. The slight discrepancies can result from the fact that a slightly updated version of the displacement curve is applied here, as well as Jaksland applying the lowest find occurence for the shoreline dating, which is slightly off from the lowest elevation of the site limits employed here. Jaksland's dates are a given a 200 and 50 year uncertainty range starting from what he gave as the earliest possible date. The sites are located in such a constrained area of the landscape that the difference in site elevation would arguably be enough to constrain the uncertainty range so that they don't overlap between sites -- information that is not integrated here. The steepness of the curve does nonetheless provide very precise dates, the accuracy of which are of course ultimately dependent on the veracity of the geological reconstruction.
